Data Specialization for a Staged Lambda Calculus

### 2015-02-28 paper/overview.tex
remove accidental page break
tightening 2.1

\pagebreak
We can attempt to improve on this algorithm by factoring out computations shared
between these calls to \texttt{qs}. In particular, the comparisons performed by
\texttt{partition} depend only on \texttt{l}, although different calls to
\texttt{partition} are made by each call to \texttt{qs}.

@@

The idea behind staged programming is to use staging annotations or
types to indicate the stage of each subterm.  To apply this
technique to our example, suppose that we have a language with two
worlds representing the two stages and a third, {\em ground} world
that contains terms with no staging.

We can write a staged version of \texttt{qs}, called \texttt{qss},
whose code is shown in \ref{fig:qs-staged}, where the first-stage code
are typeset in red, and the second-stage code are typeset in blue.  To
obtain \texttt{qss} from \texttt{qs}, we first make the input list 
first-stage and the rank second-stage.
%
More precisely the input list has type $\curr\mathrm{list}$ (an
integer list ``now''), the rank has type $\fut\rmint$ (an integer in
the ``future''), and the return has type $\fut\rmint$.
%
We can then obtain \texttt{qss} by annotating the body of \texttt{qs}
with the staging annotations \texttt{prev} and \texttt{next}, which
transition between first-stage and second-stage code, and~$\texttt{g}$, 
which marks ground terms.  
%
We also use a third function
\texttt{hold}, which can be implemented with \texttt{prev} and
\texttt{next}, for promoting a first-stage integer to a second-stage integer;
the signature of \texttt{hold} is
% 
\lstinline{1`hold : ^int -> $`2`int`}.

Our type system ensures that the staging annotations are consistent,
in the sense that computations marked as first-stage cannot depend on
ones marked as second-stage.
%
The process of adding staging annotations to unstaged code has been
the subject of extensive research (\ref{sec:related}). We assume that
these annotations have been provided; we do not consider the problem
of generating them. In this example, there are other ways to annotate
\texttt{qs}, but we chose annotations that maximize the work performed
in the first stage.

>>>>

We can attempt to improve on this algorithm by factoring out computations shared
between these calls to \texttt{qs}. In particular, the comparisons performed by
\texttt{partition} depend only on \texttt{l}, although different calls to
\texttt{partition} are made by each call to \texttt{qs}.

@@

The idea behind staged programming is to use staging annotations---in our case,
guided by types---to indicate the stage of each subterm. We consider a
language \lang\ with two modal worlds representing the two stages, and a third
{\em ground} world of unstaged terms.

In \ref{fig:qs-staged} we show a staged version of \texttt{qs}, called
\texttt{qss}, where first-stage code is colored red, and second-stage blue. In
\texttt{qss}, we regard the input list as arriving in the first stage
(with type $\curr\mathrm{list}$, a list ``now''), the input rank as arriving in
the second stage (with type $\fut\rmint$, an integer in the ``future''), and the
result as being produced in the second stage (with type $\fut\rmint$).

\texttt{qss} is obtained from \texttt{qs} by wrapping certain computations with
\texttt{prev} and \texttt{next}, signaling transitions between first- and
second-stage code, and $\texttt{g}$, signaling ground terms.  
%
We also use a function \texttt{hold},
\lstinline{1`hold : ^int -> $`2`int`},
to promote first-stage integers to second-stage integers.
%
Our type system ensures that the staging annotations in \texttt{qss} are
consistent, in the sense that computations marked as first-stage cannot depend
on ones marked as second-stage.

The process of adding staging annotations to unstaged code has been
the subject of extensive research (\ref{sec:related}). In this paper, we do not
consider this problem, instead assuming that the annotations have been provided.
In the case of \texttt{qss}, we have specifically chosen annotations which
maximize the work performed in the first stage.
